\Large \bf One-Step, Three-Factor Authentication With Custom-Fit, In-Ear EEG

### 2017-02-11 paper.tex
Updating parts of intro, tasks, collection protocol, and data validation.

In this study we undertake, to the best of our knowledge, the first
ever study of one-step three-factor authentication. In computer security,
authenticators are classified into three types: knowledge factors (e.g., passwords
and PINs), possession factors (e.g., physical tokens, ATM cards), and inherence
factors (e.g., fingerprints and other biometrics). Because three-factor
authentication (3FA) requires the user to submit one distinct instance of each
type of authenticator, it represents the strongest level of authentication security
possible.

@@

We investigate custom-fit ear-EEG technology as the platform for
investigating the feasibility, performance, and usability of one-step three-factor
authentication. In addition to the dual knowledge and inherence factors in
in previous work, this work includes a posession factor
in the form of the EEG-sensing ear-piece(s) that are custom-fitted to and worn in
their ear. These earpieces can serve as physical tokens in the same way as bank
ATM cards and wearable hardware tokens. Because the earpieces
are custom-fitted to each individual, they will likely not be able to produce good
electrical impedances when worn by a different individual.

@@

\cite{curranpassthoughts} was the first to merge in-ear EEG with passthought authentication.

@@

Custom ear impressions were made by injecting silicon into the ear canal. A cotton ball with a string attached is placed into the ear canal first.  When the silicon dries after a few minutes, the string is pulled to remove the impression from the ear canal. This impression is then scanned with a 3d scanner and the resulting scan modified based on some heuristics to achieve a comfortable fit and to ensure the intended electrode sites will make good contact with the skin. Channels are created in the 3d model to allow wire leads and associated EEG electrodes as well as a plastic tube to deliver audio. This 3d model is then sent to a 3d printer. Following this wires leads and associated AGCL electrodes are installed.

@@

7 male participants, 5 students and 2 non-students, completed our study protocol approved by the UC Berkeley Committee for Protection of Human Subjects (CPHS). The participant pool was limited due to the necessity of creating a pair of custom-fitted earpieces for each. There were three study visits per participant, the first to obtain 3D molds of participants' ears to use in creating the earpieces, the second to do a fit and impedance check once earpieces were manufactured, and a third to collect data using the earpieces while participants performed a set of mental tasks to be used in authentication analysis. Informed consent was obtained prior to study procedures. On the third study visit participants completed a short demographics questionnaire, a set up period with the openBCI and earpieces including a second impedance check, performed 9 mental tasks presented on a laptop, and finally a post-experiment questionnaire. Approximately two weeks after the third study visit, a subset of participants reported via e-mail whether or not they remembered the passthoughts they chose and performed.

@@

We used an expanded set of mental tasks with properties based on findings in related work regarding the relative strengths of different mental tasks in authentication accuracy and usability as reported by participants. Furthermore, given the in-ear placement of the electrodes and therefore the proximity to the temporal lobes containing the auditory cortex, we tested several novel authentication tasks based specifically on auditory imagery or stimuli. The nine authentication tasks and their attributes are listed in Table \ref{tab:tasks}. We purposefully selected tasks that captured a diversity of possibilities across the dimensions of external stimuli, involving a personalized secret, eyes open or closed (due to known significant effects on EEG), and different types of imagery.

\begin{table*}[t]
\centering
%% \begin{tabular}{llllll}
\begin{tabularx}{\textwidth}{llllll}
Task & Description & Stimuli? & Secret? & Eyes & Imagery\\
\hline
Breathe & Relaxed breathing with eyes closed & No & No & Closed & None\\
Breathe - Open & Relaxed breathing with eyes open & No & No & Open & None\\
Sport & Sport-related motor imagery & No & Yes & Closed & Motor\\
Song & Imagining hearing a song & No & Yes & Closed & Aural\\
Song - Open & Song task, with eyes open & No & Yes & Open & Aural\\
Speech & Imagining a spoken phrase & No & Yes & Closed & Aural\\
Listen - Tone & Listening to noise modulated at 40 Hz & Yes & No & Closed & None\\
Face & Imagine a person's face & No & Yes & Closed & Visual\\
Sequence & Imagine face, a number, and word on timed cues & Yes & Yes & Open & Visual\\
\hline
\end{tabularx}
\caption{Properties of authentication tasks. We selected tasks with a variety of different properties, but preferred tasks that did not require external stimuli, as the need to present such stimuli at authentication time could present challenges for usability, and user security.}
\label{tab:tasks}%
\end{table*}

\subsection{Data Collection Protocol}
\label{sec:org2041857}
The data collection visit took approximately 90 minutes for set up and experiment execution. The OpenBCI system we used
allows for 8 channels of simultaneous recording, along with separate ground and reference channels.
Data was initially collected with the ground placed at the center of the forehead, and using the left
mastoid as reference, though we later re-referenced to the right mastoid channel, so that our analysis reflects data from a single earpiece. Each earpiece (shown in the image below) contains three channels: one placed on the helix, and two inside the canal - one front-facing and the other back-facing. The remaining
two channels were placed on the right mastoid for later re-referencing, and at approximately Fp1 (on the 
forehead above the left eye) for validating the data collected in the ears against a scalp-based measure. 
Before beginning the experiment, the data from each channel was visually inspected and participants were asked to
blink and clench their jaws to confirm that all channels were active and properly connected. Audio stimuli was delivered through small tubes in the earpieces opening into the ear canal.

@@

In this section, we establish that the data we collected were of a neural origin, and had relatively little noise. 
Using data from two participants, we were able to confirm the custom-fit earpieces are able to collect EEG
data using three tests: good impedances measured for the ear electrodes, and alpha-band activity attenuation when a
participant's eyes were open versus closed.

@@

For the alpha-attenuation test, data from the "Breathe" task was compared with that of the "Breathe - Open" task. It is a well-
known feature of EEG data that activity in the alpha-band (approximately 8-12 Hz range) increases when the eyes are closed
compared with a similar state with eyes open. For both of our pilot participants this attenuation is clearly visible even in just
a single trial's data. To further validate, we also performed this calculation on the data collected from the Fp1 electrode and see
the effect clearly here as well. It is important to note that the left ear results are reported using the right mastoid as reference, and
the right ear results in turn using the left mastoid as reference. When using the same side mastoid for reference the effect is not 
visible, though it may be if we average across many trials. This is not surprising, as the further a reference electrode is from the active
channel the less "real" signal is being subtracted from the active channel. This has important design implications for eventual 
real-world deployment of this authentication method however, as it will likely require pieces worn on or around both ears to properly 
function, and not just one. The figures below show the alpha attenuation in the left and right ear channels, as well as Fp1.

>>>>

In this study we undertake, to the best of our knowledge, the first
ever study of one-step three-factor authentication. In computer security,
authenticators are classified into three types: knowledge factors (e.g., passwords
and PINs), possession factors (e.g., physical tokens, ATM cards), and inherence
factors (e.g., fingerprints and other biometrics). Because three-factor
authentication (3FA) requires the user to submit one distinct instance of each
type of authenticator, it represents the strongest level of authentication security
possible. The system we propose and test here uses the choice of a mental task
or "passthought" to perform as knowledge, the uniqueness of an individual's brain
activity as measured by EEG as inherence, and the physical token of custom-fit 
earpieces which could easily contain a hardware key-pair as a possession factor.
The ability to utilize all three of these security factors in a single step by performing 
a mental task of a few seconds is a promising in pursuit of extremely strong security 
while maintaining a low amount of effort and obtrusiveness to the user.

@@

We investigate custom-fit ear-EEG technology as the platform for
investigating the feasibility, performance, and usability of one-step three-factor
authentication. In addition to the dual knowledge and inherence factors in
in previous work, this work includes a possession factor
in the form of the EEG-sensing ear-piece(s) that are custom-fitted to and worn in
their ear. These earpieces can serve as physical tokens in the same way as bank
ATM cards and wearable hardware tokens by implementing a hardware key-pair.
Furthermore, because the earpieces are custom-fitted to each individual, we predicted 
they would likely not be able to produce good electrical impedances when worn by a different individual.

@@

\cite{curranpassthoughts} was the first to merge in-ear EEG with passthought authentication,
 using a single modified consumer grade EEG device among participants and achieving approximately 80 percent authentication accuracy.

@@

Custom ear impressions were made by cleaning and injecting silicon into the ear canal. A cotton ball with a string attached is placed into the ear canal first.  When the silicon dries after a few minutes, the string is pulled to remove the impression from the ear canal. This impression is then scanned with a 3D scanner and the resulting scan modified based on some heuristics to achieve a comfortable fit and to ensure the intended electrode sites will make good contact with the skin. Channels are created in the 3d model to allow wire leads and associated EEG electrodes as well as a plastic tube to deliver audio. This 3d model is then sent to a 3d printer. Following this wires leads and associated AGCL electrodes are installed.

\underline{Is there specifics on how the exactly positions of the electrodes on these pieces were chosen? If not, can simply mention 2 in the ear canal and 1 on the outer ear (helix) }

@@

7 male participants, 5 students and 2 non-students, completed our study protocol approved by the UC Berkeley Committee for Protection of Human Subjects (CPHS). Three study visits took place per participant, in the first the 3D molds of participants' ears to use in creating the earpieces were obtained, the second to do a fit and electrical impedance check once earpieces were manufactured, and a third to collect data using the earpieces while participants performed a set of mental tasks to be used in authentication analysis. Informed consent was obtained prior to study procedures. The third study visit consisted of participants completing a short demographics questionnaire, a set up period with the openBCI system and earpieces including a second impedance check, the performance of 9 mental tasks presented on a laptop, and finally a post-experiment questionnaire. Approximately two weeks after the third study visit, a subset of participants reported via e-mail whether or not they remembered the passthoughts they chose and performed.

@@

We used a set of mental tasks based on findings in related work regarding the relative strengths of different mental tasks in authentication accuracy and usability as reported by participants. Furthermore, given the in-ear placement of the electrodes and therefore the proximity to the temporal lobes containing the auditory cortex, we tested several novel authentication tasks based specifically on auditory imagery or stimuli. The 9 authentication tasks and their attributes are listed in Table \ref{tab:tasks}. Our strategy was to select tasks that captured a diversity of possibilities across the dimensions of external stimuli, involving a personal secret, eyes open or closed (due to known significant effects on EEG), and different types of mental imagery.

\begin{table*}[t]
\centering
%% \begin{tabular}{llllll}
\begin{tabularx}{\textwidth}{llllll}
Task & Description & Stimuli? & Secret? & Eyes & Imagery\\
\hline
Breathe & Relaxed breathing with eyes closed & No & No & Closed & None\\
Breathe - Open & Relaxed breathing with eyes open & No & No & Open & None\\
Sport & Sport-related motor imagery & No & Yes & Closed & Motor\\
Song & Imagining hearing a song & No & Yes & Closed & Aural\\
Song - Open & Song task, with eyes open & No & Yes & Open & Aural\\
Speech & Imagining a spoken phrase & No & Yes & Closed & Aural\\
Listen - Noise & Listening to noise modulated at 40 Hz & Yes & No & Closed & None\\
Face & Imagine a person's face & No & Yes & Closed & Visual\\
Sequence & Imagine face, a number, and word on timed cues & Yes & Yes & Open & Visual\\
\hline
\end{tabularx}
\caption{Properties of authentication tasks. We selected tasks with a variety of different properties, but preferred tasks that did not require external stimuli, as the need to present such stimuli at authentication time could present challenges for usability and user security.}
\label{tab:tasks}%
\end{table*}

\subsection{Data Collection Protocol}
\label{sec:org2041857}
The data collection visit took approximately 90 minutes for set up and experiment execution. All collection sites were cleaned with ethanol prior to electrode placement, and a small amount of conductive gel was used on each electrode. For EEG recording we used OpenBCI \cite{OpenBCI}, an open-source biosensing system to maintain an overall low cost our set up. OpenBCI has a price tag of about \$600, and thus is an affordable alternative to medical-grade EEG systems that can cost tens of thousands of dollars. Recent work has demonstrated the robustness of OpenBCI compared to an industry standard medical-grade EEG system, particularly for non-medical use cases such as ours \cite{Frey2016}. The OpenBCI system we used allows for 8 channels of simultaneous recording, along with separate ground and reference channels. Data was initially collected with the ground placed at the center of the forehead (a common approach to not bias one side of the head over the other), and using the left mastoid as reference, though we later re-referenced to the right mastoid channel so that our analysis reflects data from a single earpiece. Each earpiece (shown in the image below) contains three channels: one placed on the helix, and two inside the canal - one on the anterior canal wall (front-facing) and the other on the posterior wall (back-facing). The remaining two channels, AGCL ring electrodes, were placed on the right mastoid for later re-referencing, and at approximately Fp1 (the forehead above the left eye) for validating the data collected in the ears against a scalp-based placement. Before beginning the experiment, the data from each channel was visually inspected using the OpenBCI GUI and participants were asked to blink and clench their jaws to confirm that all channels were active and properly connected. Audio stimuli were delivered through small tubes in the earpieces opening into the ear canal.

@@

In this section, we establish that the data we collected were of a neural origin, and had relatively little noise. We were able to confirm the custom-fit earpieces are able to collect EEG data using two tests: good impedances measured for the ear electrodes, and alpha-band activity attenuation when a participant's eyes were open versus closed.

The first quality check was examining the electrical impedances achieved for the ear pieces as a low impedance (<10 kOhm) 
is best for obtaining quality EEG signals. Table \ref{tab:impedances} below summarizes the impedances across the seven participants' six ear channels. With the exception of a few channels in select participants, impedances achieved were overall very good.

\begin{table}[h]
\begin{center}
\begin{tabular}{lrrrrrr}
& \multicolumn{6}{c}{Impedances [kOhm]} \\
\cline{2-7}
& \multicolumn{3}{|c|}{Left ear} & \multicolumn{3}{c|}{Right ear} \\
Subject & Helix & Front & Back & Helix & Front & Back \\
\hline
1 & 4 & 4 & 4 & \textless1 & 4 & 3\\
2 & 9 & 5 & 4 & 3 & 4 & 4\\
3 & 4 & 5 & 4 & 9 & 6 & 9\\
4 & 4 & 5 & 4 & 3 & 16 & 9\\
5 & 9 & 20 & 7 & 3 & 7 & 9\\
6 & 5 & 8 & 2 & 1 & 1 & 9\\
7 & 2 & 9 & 8 & 7 & 5 & 6\\
\end{tabular}
\end{center}
\caption{Electrical impedances measured for earpiece electrodes.}
\label{tab:impedances}
\end{table}

@@

For the alpha-attenuation test, data from the "Breathe" task was compared with that of the "Breathe - Open" task. It is a well-
known feature of EEG data that activity in the alpha-band (approximately 8-12 Hz range) increases when the eyes are closed
compared with a similar state with eyes open. For our participants, this attenuation is clearly visible even in just
a single trial's data. To further validate, we also performed this calculation on the data collected from the Fp1 electrode and see
the effect clearly here as well. It is important to note that the left ear results are reported using the right mastoid as reference, and
the right ear results in turn using the left mastoid as reference. When using the same side mastoid for reference the effect is not 
visible, though it may be if we average across many trials. This is not surprising, as the further a reference electrode is from the active
channel the less "real" signal is being subtracted from the active channel. This has important design implications for eventual 
real-world deployment of this authentication method however, as it will likely require pieces worn on or around both ears to properly 
function, and not just one. The figures below show the alpha attenuation in the left and right ear channels, as well as Fp1 of one participant for example.
